

> Human personalities are rather like fractals. It is not just that what we do in the large-scale narratives of our lives—love, career, friendships—tends to be somewhat consistent over time, with us often repeating the same kinds of triumph or mistakes. Rather, what we do in tiny interactions like the way we shop, dress or talk to a stranger on a train or decorate our houses, shows the same kinds of patterns as can be observed from examining a whole life.

> -- <cite>Daniel Nettle, [Personality: What Makes You the Way You Are](http://www.columbia.edu/cgi-bin/cul/resolve?clio14146523)</cite>



This handbook is intended to serve as a resource allowing you to see from the outset what lies ahead, and as a place to collect your notes as we progress through the course.

It is important that you know in advance what to expect from this lab so that you can plan your time accordingly. That information is all here in this handbook. This laboratory handbook outlines the procedures, rationales, and techniques for lab exercises, and provides an overview of the accompanying assignment for each week. 

You’ll find a section for each lab session, giving a brief overview of the main themes and questions for that session, some suggested (optional) additional readings, and information about the accompanying assignment. 

Each lab section contains blank space for recording notes and for collecting observations. Therefore, you should bring the lab manual to each meeting of the lab.

You’ll also find a number of appendices, covering things like searching the academic literature, how to approach reading journal articles, summaries of the basic statistics you should be familiar with, examples of informed consent and debriefing forms, and an extensive guide to writing a lab report.

We will reiterate much of this information as part of each lab session's instructional component, and we'll greatly expand on the themes through the activities you'll do and our collective discussions that follow. 

You should skim through this handbook at the beginning of the course so that you have at least a rough sense of what will be required and when, and so you know where you can find relevant information as and when you need it.

## Syllabus {.unnumbered}

### Instructor

Dr. Rob Brotherton (<rbrother@barnard.edu>)

Office hour: TBC

### Time & venue

```{r}
#| echo: false

details <- tibble::tribble(
  ~Section, ~Time, ~Venue,
  "001",    "Monday 10:10--1PM", "Milbank 410",
  "002",    "Monday 1:10--4PM", "Milbank 410"
)

knitr::kable(details)
```


### Class schedule

```{r}
#| echo: false

schedule <- tibble::tribble(
  ~Date, ~Topic, ~Assignment,
  "9/11",   "Course overview",           "",
  "9/18",   "Project planning",          "",
  "9/25",   "Data cleaning",             "",
  "10/2",   "Analysis",                  "Presentation slide",
  "10/9",   "Presentations",             "Project 1 report",
  "10/16",  "Project planning",          "",
  "10/23",  "Data cleaning & analysis",  "",
  "10/30",  "Visualization & reporting", "Presentation slide",
  "11/6",   "Presentations",             "Project 2 report",
  "11/13",  "Scale planning",            "",
  "11/20",  "Questionnaire design",      "",
  "11/27",  "No class (Thanksgiving)",   "",
  "12/4",   "Analysis & reporting",      "Presentation slide",
  "12/11",  "Presentations",             "Project 3 report"
)

knitr::kable(schedule) |>
  kableExtra::row_spec(12, italic = TRUE, color = "grey") |>
  kableExtra::pack_rows("Project 1: Correlation", 2, 5, indent = FALSE) |>
  kableExtra::pack_rows("Project 2: Multiple regression", 6, 9, indent = FALSE) |>
  kableExtra::pack_rows("Project 3: Scale design", 10, 14, indent = FALSE) |>
  kableExtra::add_footnote("Assignments due by the following class.", notation = "symbol") |>
  kableExtra::kable_styling(full_width = TRUE)

```

## Course overview

This lab is designed to offer experience participating in, reading, critiquing, and proposing research addressing social psychological phenomena. The lab will usually be taken concurrently with PSYC-BC2138 Social Psychology Lecture, and will expand upon the methodological and analytic issues introduced there, as well as giving you the opportunity to explore in greater depth a topic of your choosing from within (or beyond) those covered in the lectures by proposing your own original research project.


### Research projects

The semester is broken into 3 projects. Each section will culminate in an in-class presentation and submission of a brief write up of your project.



### Class format & participation

Labs are substantially more interactive and discussion-based than the traditional lecture format, and depend on everyone’s participation. Your active participation across the semester will therefore contribute a substantial portion of your grade. 

If you have questions, thoughts, or ideas you want to share, feel free to do so at any time (while keeping within the bounds of polite conversation, obviously--don't interrupt or talk over other people! But do feel free to respond to others without having to raise your hand or wait to be called on). Everyone will get the most out of this lab when the discussion can develop organically and everyone feels free to be part of the conversation if & when they have something to add.

Being part of the in-person discussion is one obvious way to participate, but it’s not the only way. Different people have different styles of participation, and the lab is designed to try and accommodate and encourage different approaches (as well as allowing for the greater-than-usual possibility of missing class for health reasons). Your level of engagement with your project partner(s), your TA and Prof. Brotherton as you work through your project is also an important form of participation. You can also participate by coming to office hours.

At a minimum, I’ll be looking for some form of participation (loosely defined) from you every week; higher participation grades will be earned through regular, enthusiastic, productive participation (note that quality is more important than quantity).



### Workload

As a general rule for the amount of time students should expect to commit to classes, the college suggests three hours per week in or outside of class per credit. Since this class is worth 1.5 credits, that corresponds to 4.5 hours per week, split between time in the classroom and time spent completing the associated assignments.


### Final Grades

Your final letter grade for the course is a product of your scores for each assignment, weighted as indicated in the class schedule, and determined according to the following boundaries:

    Letter grade:  A+ A  A-	B+ B  B- C+ C  C- D  F
    Numeric score: 97 93 90 87 83 80 77 73 70 60 <60




## Course policies

### Attendance & timeliness

In-person attendance of each lab session is expected, except in emergencies. Normally it is departmental policy to remove students who miss more than two sessions from the course; however, given ongoing revisions to college-wide health-related policies, exceptions may be made. If you are feeling unwell, you should not come to class and notify me of nonattendance before class if possible. When you are attending, please arrive on time for class. Frequent lateness will impact your participation grade.

### Assignment deadlines & late policy

Assignments are listed in the class schedule next to the class session to which they correspond. Most assignments are due at 5pm the day after class. All assignments relate to your project proposal—the idea is to build this major piece of work incrementally over the course of the semester. The 5pm next-day deadline is necessary to allow time to provide substantive feedback to everyone in advance of the next assignment.

A grade penalty of 5 points will be applied for each day (or part thereof) that an assignment is late (up to a maximum of 6 days; work not submitted before the next lab will receive a score of zero). For example, if your work is A+ quality but is submitted a day-and-a-half late, you will only receive a B+. Due to time constraints, you may also not receive feedback on late work. 

This policy is intended to incentivize timely submission while easing the stress of genuine emergencies. When things come up that prevent timely submission you can prioritize accordingly, knowing that a small penalty on one assignment for this lab will not tank your final grade.

### Academic integrity

Students are expected to follow the Barnard Honor Code, available at <https://barnard.edu/honor-code>.

Note that even though you may collaborate in lab and on the research proposal, you may not collaborate on written assignments. You must write your own reports. 

### Academic accommodations and general wellness

It is always important to recognize the different pressures, burdens, and stressors you may be facing, whether personal, emotional, physical, financial, mental, or academic. The faculty and administration recognize this, and are prepared to provide assistance to students in need. I encourage you to seek advice from your advisor, Dean, the [Center for Accessibility Resources & Disability Services (CARDS)](https://barnard.edu/disability-services), or [Barnard Health & Wellness](https://barnard.edu/health-wellness) as needed. Please let me know of any issues you wish to share with me that you feel are impacting your ability to complete the course to the best of your ability. Though it isn’t always easy, it is better to proactively seek help rather than letting problems build up.


<style>
#title-block-header {
  text-align: center;
}

#title-block-header.quarto-title-block.default .quarto-title-meta {
  display: unset;
}
</style>

