---
execute: 
  eval: false
---

# Lab 11: Data preparation & analysis

Groups should have decided on what variables to use. By the end of this session you should have prepared your data and completed your analysis.


### Data wrangling tips

Since everyone will be doing different things, Professor Brotherton work with groups individually to provide guidance on data preparation and analysis where help is needed. However, here are some hints about how to achieve common tasks.

#### Compute an average of several variables.

Some psychological measures consist of more than one question, and you need to compute the average of each participant's answers to all the relevant questions. `mutate()` and `rowMeans()` can be used.

```{r}
my_data |> 
  mutate(mean = rowMeans(across(all_of(scale_vars))))
```


#### Compute a sum score.

If you need to add scores across several questions, `rowSums()` can be used.

```{r}
my_data |> 
  mutate(sum = rowSums(across(all_of(scale_vars))))
```


#### Ntiles.

Occasionally you might like to split a continuous measure into "ntiles," meaning a number of roughly equally-sized groups. The `ntile()` function can be used to this.

```{r}
my_data |> 
  mutate(income_bracket = ntile(income, 2))
```


#### Arbitrary groups

```{r}
my_data |> 
  mutate(age_group = case_when(age < 30 ~ "Young", age >= 30 ~ "Old"))
```



<!-- ### Regression with interaction -->

<!-- When specifying a regression model which includes the interaction between two predictors, the only difference is that you separate the names of the predictor variables with a `*` rather than `+`. -->

<!-- ```{r} -->
<!-- model <- lm(feeling_thermometer ~ conscientiousness_mean * agreeableness_mean, data = my_data_complete) -->

<!-- summary(model) -->
<!-- ``` -->

<!-- If the interaction term is significant, it's also important to conduct a post hoc analysis to probe the interaction. This often involves plotting the interaction or calculating the effect of one variable at different levels of the other variable, to get a clearer understanding of how the predictors interact to affect the outcome. -->

<!-- Keep in mind that interpreting interaction terms can be complex. It is crucial to consider the nature of your variables, the context of your research, and the practical significance of the interactions, not just the statistical significance. -->
